---
title: "discordance"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ape)
library(reshape2)
library(dplyr)
library(phylogram)
library(dendextend)
library(pbapply)
```

Download all files from microreact. Not all the GPSC seem to have a microreact project associated. the subset below all had files.

```{r}
gpcs.test<-c(1:27,30:34, 37:41, 43, 47:48, 50:58, 61:62,67:68,70,72, 76,78, 79,80,81,90,91,93, 94,97,103,105,117, 131)
```

```{r, eval=F}
library(httr)
library(phylogram)
library(dendextend)
for(i in gpcs.test){
path <- paste0("https://microreact.org/api/viewer/tree?project=gpsGPSC",i)
path2<-paste0("https://microreact.org/api/viewer/data?project=gpsGPSC",i)
download.file(path2, destfile=paste0("./metadata/data_gpcs",i,".csv"))
download.file(path, destfile=paste0("./trees/tree_gpcs",i,".nwk"))
}
```


### Read in all trees and save in a list
```{r}
trees <- vector("list",length(gpcs.test))
trees.to.cut <- vector("list",length(gpcs.test))

for(i in 1:length(gpcs.test)){
  trees[[i]]<-ape::read.tree(paste0("microreact/tree_gpcs",gpcs[i],".nwk"))
  trees.to.cut[[i]]<-read.dendrogram(paste0("microreact/tree_gpcs",gpcs[i],".nwk"))

}  
```

###Import the metadata
```{r}
metads <- vector("list",length(gpcs.test))
for(i in 1:length(gpcs.test)){
  metads[[i]]<-read.csv(paste0("microreact/data_gpcs",gpcs[i],".csv"))
} 
```

## classify the samples based on a cutpoint
```{r}
tree.cutter<-function(tree.list, cutpoint){
  test1<-as_hclust_fixed(tree.list)
  test2<-cutree(test1, h=cutpoint)
  test3<-as.data.frame(test2)
  names(test3)<-'tree.grp'
  test3$ID<-row.names(test3)
  return(test3)
}

tree.grps200<-lapply(trees.to.cut,tree.cutter,cutpoint=200)

```

## Merge in with meta data
```{r}
metads<-lapply(metads, function(x){
  x$ID<-as.character(x$ID)
  return(x)
 })
meta.merge<-function(tree.ds, meta.ds){
  test1<-merge(tree.ds, meta.ds, by='ID')
  test1$phylo.grp<-paste(test1$GPSC, test1$tree.grp, sep='_')
  test1$In_Silico_Serotype<-as.character(test1$In_Silico_Serotype)
  test1<-test1[,c('ID', 'In_Silico_Serotype','phylo.grp','In_Silico_Serotype' )]
  return(test1)
}

grp.assign200<-mapply(meta.merge,tree.ds=tree.grps200, meta.ds=metads , SIMPLIFY=F)
#phylo.grps200<- do.call('rbind.data.frame',grp.assign200)
```


Distance matrix
Generate a distance matrix based on the phylogenetic tree. Then identify the closest neighbors of switched variants on same part of the tree (based on the tree.cut)
```{r}
distance_mat_fun <- function(tree.ds, meta.ds){
  
  # for(i in 1:length(trees)){
  #   
  # tree.ds=trees[[i]]
  # meta.ds=grp.assign200[[i]]
  # print(i)
  
  dist.t1 <- cophenetic.phylo(tree.ds)
  dist.t1.m <- reshape2::melt(dist.t1)
  
  dist.t1.m <- merge(dist.t1.m, meta.ds[,c('ID','In_Silico_Serotype',"phylo.grp")], by.x='Var1', by.y='ID', all.x=T)
  dist.t1.m <- merge(dist.t1.m, meta.ds[,c('ID','In_Silico_Serotype',"phylo.grp")], by.x='Var2', by.y='ID', all.x=T)
  names(dist.t1.m) <- c('ID2','ID1','distance','sero1','phylo_grp1', 'sero2','phylo_grp2')
  
  dist.t1.m <- dist.t1.m[dist.t1.m$phylo_grp1 == dist.t1.m$phylo_grp2 , ] #only consider pairs on same part of the tree
  
  dist.t1.m$change_st <- 1*(dist.t1.m$sero1!=dist.t1.m$sero2)
  
  dist.t1.m <- dist.t1.m[dist.t1.m$distance>0,] #remove diagonal from the distance matrix
  
 # hist(dist.t1.m$distance)
  
  
  #dist2 <- dist.t1.m[dist.t1.m$distance<200,] #restrict to more closely-related strains
  
  # ave_dist <- dist2 %>%
  #   group_by(change_st) %>%
  #   summarize(ave_dist=mean(distance))
  
  if(sum(dist.t1.m$change_st)>0){
  closest_match <- dist.t1.m[dist.t1.m$change_st==1,] %>%
    group_by(sero1, sero2) %>%
    arrange(distance, .by_group = TRUE) %>%
    mutate(distance_rank = 1:n())

  }
  
return(closest_match[closest_match$distance_rank==1,])
  }else{
    return(NULL)
  }
}

all_switches <- pbmapply(FUN=distance_mat_fun, tree.ds=trees,meta.ds=grp.assign200, SIMPLIFY = F)
```



